---
title: "Project proposal"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project proposal:
Our project seeks to use a variety of Gapminder country level statistics from 2016 to determine which variables correlate strongly with the overall level of happiness in a country (as measured by the Happiness Index), and how.

```{r}
library(tidyverse)
```


```{r}
df=readRDS("./merged_gapminder_happiness.rds")
df=df%>%select(-forest,-mortality) # forest currently contains only NA values
```


```{r}
summary(df)
colnames = colnames(df)
colnames = colnames[-c(1, 2)]
for(i in colnames) {
  numeric = sapply(na.omit(df[, i]), function(x){
    as.numeric(x)
  })
  hist = hist(numeric, breaks = 20, main = paste("Distribution of", i))
}
```

Description & distribution of each variable:
                         For each country:
parliament: Percentage of national parliamentary seats held by women. Lower and upper houses combined.
Distribution - skewed right, normal.

hapiscore_whr: This is the national average response to the question of life evaluations asking the following “Please imagine a ladder, with steps numbered from 0 at the bottom to 10 at the top. The top of the ladder represents the best possible life for you and the bottom of the ladder represents the worst possible life for you. On which step of the ladder would you say you personally feel you stand at this time?” This measure is also referred to as Cantril life ladder. Gapminder has converted this indicator's scale from 0 to 100 to easily communicate it in terms of percentage.
Distribution - mild hump/stretched-out normal curve with wide spread

aged_15plus_unemployment_rate_percent: Percentage of total population, age group above 15, that has been registered as unemployed during the given year.
Distribution - normal

right: Fundamental Rights in the form of liberal and social rights support both fair representation and the vertical 
mechanism of accountability that the first attribute seeks to achieve. This attribute is composed of three 
subattributes: access to justice, civil liberties, and social rights and equality. The three subattributes were 
aggregated into the Fundamental Rights index using BFA (source: Skaning 2019 via https://www.idea.int/gsod-indices/sites/default/files/idea-gsodi-2019-codebook-v3.pdf )
Distribution: normal/hump, wide spread

sanitation: percentage of people using improved sanitation facilities that are not shared with other households.
distribution - uniform/linear then with sharp jump/exponential increase at 100%.

hdiindex: Human Development Index (HDI) https://www.hdr.undp.org/en/content/human-development-index-hdi
distribution - concave down increasing, wide spread.

edu: Education index calculated based on Avg years of schooling, taking values 0 as minimum and 15 as maximum
Distribution - wide spread.

internetusers: Percentage of Population with stable Internet access
Distribution - multimodal, wide spread.

sdi: The Sustainable Development Index is an efficiency metric, designed to assess the ecological efficiency of nations in delivering human development. It is calculated as the quotient of two figures: (1) a “development index” based on the Human Development Index, calculated as the geometric mean of the life expectancy index, the education index, and a modified income index; and (2) an “ecological impact index” calculated as the extent to which consumption-based CO2 emissions and material footprint exceed per-capita shares of planetary boundaries.
Distribution - seemingly linearly increasing, or skewed left.

forestcoverage: Percentage of total land area that has been covered with forest
Distribution - Seemingly decreasing.

militaryexpenditure: Military expenditure as a percentage of GDP.
Distribution - Right skewed normal with outliers at larger values. 

mortality: Newborn mortality rate per 1000
Distribution - decreasing

wateraccess: Percentage of people with at least basic water services
Distribution - exponentially increasing.

broadband: Broadband subscribers per 100 people
Distribution - steep drop/potentially exponentially decreasing.

agriculture: Percentage of land area that is arable
Distribution - wide-ranging

debt: Total debt service (% of GNI)
Distribution - Sharp decrease then flat?

gdppercapita: GDP per capita
Distribution  - decreasing with outliers

foodinsecurity: Prevalence of moderate or severe food insecurity in the population
Distribution - normal skewed right?
